---
title: "2020 Elections in Madison"
author: "Harald Kliems"
date: "11/3/2020"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
   echo=F, results=F, message=F, error=F, warning=F
)
```

```{r}
library(tidyverse)
library(googlesheets4)
library(tidycensus)
library(tmap)
library(sf)
library(readxl)
```

```{r wards}
# Ward shapefile from City of Madison OpenData portal
wards <- read_sf("data/Wards.shp")
```

# Preliminary results by ward
Pulled from the [Dane County Clerk website](https://elections.countyofdane.com/Precincts-Result/124/0004) at 6:42am on Nov 4. Wards with fewer than 25 total votes are excluded.

```{r}
results <- read_excel("data/ElectionResults.xlsx", skip = 7, 
                      col_names = c("precinct",
                                    "biden",
                                    "trump",
                                    "con",
                                    "lib",
                                    "asp",
                                    "write_in"))
```

```{r}

results <- results %>% 
  mutate(other = con + lib + asp + write_in,
  total_votes = biden + trump + other,
  biden_pct = biden/total_votes*100,
  trump_pct = trump/total_votes*100
  ) %>% 
  select(-c(con, lib, asp, write_in))
results_long <- results %>% 
  pivot_longer(cols = 2:4, names_to = "candidate", values_to = "votes") %>% 
  filter(str_detect(precinct, "C Madison")) %>% 
  mutate(precinct = as.numeric(str_sub(precinct, -3)))
```

```{r}
results_joined <- wards %>% 
  inner_join(results_long, by = c("WARD" = "precinct")) %>% 
  filter(total_votes > 25)

tmap_mode("view")
```
```{r results=T}

tm_shape(results_joined) +
  tm_polygons("biden_pct",
          style = "jenks",
          title = "Joe Biden",
          popup.vars = c("Vote share Bide (%)" = "biden_pct", "Vote share Trump (%)" = "trump_pct", "Total votes cast" = "total_votes"),
          legend.format=list(fun=function(x) paste0(formatC(x, digits=0, format="f"), "%")))
```



# Absentee ballots

Local journalist [Jason Joyce](https://twitter.com/jjoyce) compiled data on absentee ballots returned in a [neat spreadsheet](https://docs.google.com/spreadsheets/d/15cIkjmRWpipYfBNtUbVuS4_aaVEuSIXv2DbNEx5k00I/edit#gid=613969698). He last updated the data on November 1. I quickly wrote some code to create maps of his data.




```{r}
gs4_deauth()
absentee <- read_sheet("https://docs.google.com/spreadsheets/d/15cIkjmRWpipYfBNtUbVuS4_aaVEuSIXv2DbNEx5k00I/edit#gid=613969698", skip = 2, n_max = 134, col_types = "iiiinininin", col_names = c("WARD", "not_returned", "returned", "total", "perc_returned", "reg_voters", "perc_reg_voters", "2016_voters", "perc_2016_voters", "2016_absentee", "perc_2016_absentee")) %>% 
  drop_na(any_of("perc_returned"))
```



Here is the percentage of absentee ballots returned versus requested. I filtered out wards with fewer than 15 registered voters.
```{r}
absentee_sf <- wards %>% 
  inner_join(absentee, by = "WARD") %>% 
  mutate(perc_returned = round(perc_returned *100, 1),
         perc_2016_voters = round(perc_2016_voters * 100, 1)) %>% 
  select(-OBJECTID)
```


```{r results= T}

tmap_mode("view")
absentee_sf %>% 
  filter(reg_voters > 15) %>% 
tm_shape() +
  tm_polygons("perc_returned",
              style = "jenks",
              title = "Absentee ballots returned<br> as of Nov 1",
              popup.vars = c("Absentee ballots returned (%)" = "perc_returned",
                             "Ward" = "WARD",
                             "Number of absentee ballots returned" = "returned"),
              legend.format=list(fun=function(x) paste0(formatC(x, digits=0, format="f"), "%")))
```

And here are the absentee ballots returned compared to total votes cast in the 2016 election. I filtered out wards with fewer than 15 total votes in 2016, as they messed up the scale.

```{r results=T}

absentee_sf %>% 
  filter(`2016_voters` > 15) %>% 
tm_shape("WARD") +
  tm_polygons("perc_2016_voters",
              alpha = .7,
              style = "jenks",
              popup.vars = c("2020 absentee returned vs 2016 total (%)" = "perc_2016_voters",
                             "Ward" = "WARD",
                             "Number of absentee ballots returned" = "returned"),
              title = "2020 absentee ballots returned<br>compared to 2016 total votes",
              legend.format=list(fun=function(x) paste0(formatC(x, digits=0, format="f"), "%")))
```

Note the low percentages on and around the UW campus and the high percentages on the far west and far east sides. 

